\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-10-04 paper/main.tex
filled in impact and conclusions sections, updated figure, moved a bunch of old text to the boneyard

Code sharing is a core component of the open science movement
that has inspired a full ecosystem of tools and packages.  However,
sharing code, in and of itself, does not guarantee that others will be
able to \textit{reproduce} the desired results.  For eample, research
code often requires installing other software packages that extends
the language's basic functionality.  Within the Python community~\citep{vanR95},
external packages that are published in the most popular 
repositories~\citep{pypi, Anac12} are associated with version numbers
and tags that enable users to guarantee that they are installing
exactly the same code across different computing environments.
Despire that it is \textit{possible} to manually install the intended version numbers of
every dependency of a Python script or package, doing so may cause
conflicts within the user's computing environment that interferes with
the functionality of  \textit{other} code.

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python ecosystem.}}
\label{fig:code-sharing}
\end{figure}

@@

At another extreme, virtual machines~\citep{Gold74, AltiEtal05, vmware} provide a complete hardware-level
simulation of the desired system.  Virtual machines are typically
fully isolated from the user's system such that installing or running
software on a virtual machine does not impact the user's primary
operating system or computing environment.  Containers~\citep[e.g.,][]{Merk14,
  KurtEtal17} provide a similar ``isolated'' experience.  Although
containerized environments do not specify hardware-level operations,
they are typically packaged with a complete operating system, in
addition to a complete copy of Python and any relevant dependencies.
Virtual environments~\citep[e.g.,][]{vanREtal14} also provide a
computing environment that is largely separated from the user's main
environment.  They incorporate a copy of Python and the target
software's dependencies, but virtual environments do not specify or
reproducee a complete copy of the original operating system.  Each of
these systems (virtual machines, containers, and virtual environments)
guarantees (to differing degrees-- at the hardware-level, operating system
level, and Python environment, respectively) that the relevant code
will run similarly for different users.  However, each of these
systems also requires installing independent tools that
can be resource intensive or burdensome to install or configure.

We designed \texttt{davos} to occupy a ``sweet spot'' between these
extremes.  \texttt{davos} is a notebook-installable package that adds
functionality to the original.  Like standard Jupyter notebooks,
\texttt{davos}-enhanced notebooks allows researchers to include text,
executable code, and media within a single file.  No further setup or
installation is required, beyond what is needed to run standard
Jupyter notebooks.  And like virtual environments \texttt{davos}
provides a convenient mechanism for fully specifying (and installing,
as needed) a complete set of Python dependencies, including the
relevant version numbers.

% % STOPPED HERE...

% Sharing a project's code base is most valuable when that code may be
% run on

% Modern scientific research frequently entails writing software code
% for a wide variety of purposes throughout the scientific process.
% Researchers across disciplines may design and implement complex
% experiments; collect, store, and analyze large datasets; create
% visualizations for presentations and publications; and share their
% findings and techniques with peers, students, and the broader public
% through tutorials, demos, workshops, and classes.  However, one
% fundamental requirement of virtually any form of research-related code
% is that its behavior and outputs remain consistent and predictable, no
% matter when, where, or by whom it is run.  This stability can be
% crucial, for example, to ensuring that data are collected under the
% same conditions (e.g., across recordings, subjects, or physical
% locations)
% %^ or locations, e.g., in different test rooms, on different machines, at different sites, etc.
% over multiple months or years, and that they can be accessed,
% processed, and analyzed consistently by a research team that may be
% spread across multiple institutions or countries.  Additionally,
% modern open science practices encourage publicly sharing research code
% and data so that others may explore, reproduce, learn from, and build
% upon existing work.  Much of the benefit afforded by freely available
% research code depends on users' ability to execute it and achieve the
% same result as its original author.

% Python \cite{vanR95} has become one of the most widely used and
% fastest-growing scientific programming languages, in part by combining
% an accessible, high-level syntax with a rich ecosystem of powerful
% third-party tools that facilitate rapid development and collaboration
% \cite{MullEtal15}.  The Python ecosystem offers an extensive data
% science toolkit with platforms for interactive programming (e.g.,
% Project Jupyter \cite{KluyEtal16}, Google Colaboratory),
% community-maintained libraries for data manipulation (e.g., NumPy
% \cite{HarrEtal20}, SciPy \cite{VirtEtal20}, Pandas \cite{McKi10}) and
% visualization (e.g., Matplotlib \cite{Hunt07}, seaborn \cite{Wask21}),
% frameworks for training complex machine learning models (e.g.,
% scikit-learn \cite{PedrEtal11}, TensorFlow \cite{AbadEtal15}, Hugging
% Face~\cite{WolfEtal20}), and myriad other resources.  However, this
% heavy emphasis on third-party libraries also presents a challenge to
% writing and sharing stable, reproducible scientific Python code:
% different versions of the same library may behave differently, adopt
% changes in syntax, expose different functions and interfaces, add or
% drop support for specific hardware or software, write (or expect to
% read) files in different formats, fix (or introduce) bugs, and so on.
% While these issues exist to some extent in any software language or
% ecosystem, they have a particular impact on the Python community due
% to its unusually rapid growth relative to other languages.  Ensuring
% Python code behaves consistently over time and across users therefore
% typically requires ensuring it is always run with the same specific
% set of versions for each third-party package used.
% %Ensuring Python code behaves consistently over time and across users therefore typically requires guaranteeing that the same specific version of each required third-party package is used, every time it is run.

% One common approach to solving this problem is to create containerized
% or virtualized Python environments (e.g., using Docker \cite{Merk14},
% Singularity \cite{KurtEtal17}, or conda \cite{Anac12}) tailored to
% individual applications.  A researcher may build such an environment
% around a particular experiment or analysis pipeline, and exclusively
% run their code from inside it, ``entering" and ``exiting" the
% environment before and after each time they do so.  They can also
% distribute their custom environment alongside their code as a
% configuration file that explicitly lists required package versions,
% enabling others to build identical copies for themselves.  This allows
% research teams to deploy experiments on multiple machines for more
% efficient data collection, collaborate on analyses without introducing
% conflicts or inconsistencies, and publicly share their study designs
% and results for others to reproduce, replicate, or adapt to study new
% questions in the future.

% While often effective, this approach bears two notable drawbacks.
% First, it can add substantially to the technical knowledge, computing
% resources, and initial setup needed to run or share the actual code of
% interest.  For example, sharing code for an analysis or tutorial that
% relies on a particular Docker image to run properly would of course
% necessitate writing and distributing extra configuration files and
% setup instructions.  But far more burdensomely, it also requires that
% anyone who may want to run the code (in addition to the author seeking
% to share it!) first be able to install and navigate additional
% software that is likely far more complex and resource-intensive than
% the actual analysis or tutorial code it facilitates running.  This can
% introduce a need for both a degree of computer literacy and
% computational resources that may not be universally accessible,
% particularly to students or other early-career scientists hoping to
% learn from publicly available tutorials.  These added prerequisites
% clash with the simplicity and accessibility that have contributed to
% Python's popularity, and can create significant barriers to both
% contributing to and taking advantage of open science resources.

% Second, while many existing tools allow users to initially populate a
% Python environment with a fixed set of packages and package versions
% (e.g., from a \texttt{requirements.txt}, \texttt{pyproject.toml},
% \texttt{environment.yml}, \texttt{Pipfile}, \texttt{Dockerfile RUN}
% instruction, etc.), few, if any, ensure that these specified
% requirements \textit{remain} satisfied after they are first installed.
% The ability to modify an environment after its creation is useful in
% many cases (e.g., to install additional software, when
% needed). However, this also makes it easy to inadvertently alter
% existing packages, potentially leading to subtle issues with code that
% relies on them.  For instance, suppose a researcher has implemented a
% series of analyses using version 1.0 of ``Package \textit{X}," and
% decides to perform an additional analysis that requires installing
% ``Package \textit{Y}."  If Package \textit{Y} depends on version 0.9
% of Package \textit{X}, then Package \textit{X} will be downgraded to
% accommodate this new requirement, potentially altering or breaking
% previous analyses when they are rerun later, either by the researcher
% or someone with whom they've shared their code.  Further, if some
% analyses require Package \textit{Y} while others rely on features of
% Package \textit{X} not implemented until version 1.0, it's unclear
% which version the researcher should install in their environment.

% %The \texttt{davos} package provides a novel, Python-native framework for creating, sharing, and running reproducible workflows that was designed to address each of these issues.
% The \texttt{davos} package provides a novel, Python-native framework
% for creating reproducible workflows that was designed to address each
% of these issues. \texttt{davos} allows users to specify dependencies
% directly within the code that...

% \comment {
% POINTS TO MAKE/THINGS TO SAY HERE TO WRAP UP (or maybe save some for impact section?):
% - davos simplifies the process of writing, sharing, and running reproducible code (eliminates need for additional complex env management tools, syntax is simple, packages are just installed at runtime without additional setup beforehand, etc.)...
% - ...it also provides stability/additional protections against accidental changes in ways other tools don't (i.e., addresses issue 2 above) by verifying (and installing if necessary) requirements each time code is run
% - ... while simultaneously enabling new package management patterns (using multiple versions versions of package in same notebook, etc.)
% - it's lightweight and intuitive (unlike big/complex env management tools like docker, etc.)
% - "davos was specifically designed for use with Jupyter notebook environments, and can be used either on its own or in tandem with another environment management tool"
% - "\texttt{davos} was designed specifically for use in Jupyter notebook (formerly called IPython notebook \cite{PereGran07}) environments, and supports Jupyter notebooks \cite{KluyEtal16}, JupyterLab \cite{GranGrou16}, Google Colaboratory, Binder \cite{RagaWill18}, and IDE-based notebook editors."

% - maybe don't mention smuggle statement \& onion comment explicitly in this paragraph since they haven't been defined/explained yet, or if we do, then add some sort of "see \_\_\_ below" with \\ref to relevant subsection
% }

%Introduce the scientific background and the motivation for developing the software.

%Explain why the software is important, and describe the exact (scientific) problem(s) it solves.

%Indicate in what way the software has contributed (or how it will contribute in the future) to the process of scientific discovery; if available, this is to be supported by citing a research paper using the software.

%Provide a description of the experimental setting (how does the user use the software?).

%Introduce related work in literature (cite or list algorithms used, other software etc.).

@@

\subsection{Software architecture}
The \texttt{davos} package is structured as two subpackages: a set of ``core" modules that implement...

@@

% ------------------------------------------------------- IMPACT -------------------------------------------------------
\section{Impact}
\comment{
- emphasize research domain-independent application (Journal's goals/values page says they prioritize this in submissions)
- applicability not limited to research
    - regression (?) testing
    - ***^^not sure this should be included -- hurts framing of davos as "research software"***
- makes writing and running publicly available reproducible code more accessible (lowers required software comfort/familiarity level \& computing resources vs. installing docker/singularity/conda)
    - also removes these requirements from other scenarios, e.g., easier for new/less experienced students/RAs to get involved with analyses
- mention something about not having to teach environment management to teach day 1 python lesson, or spend hours debugging conda installation in order to share tutorial or demo with students
    - e.g., "researchers who teach \_\_\_ may be familiar with the experience of running a lesson or workshop that requires students to use Docker (or similar...) to run their code, only to spend the majority of class time debugging students' computing environments rather than teaching the lesson itself."
}

Since its initial release, \texttt{davos} has found use in a variety
of applications.  In addition to managing data analysis environments
for multiple ongoing research studies, \texttt{davos} is being used by
both students and instructors in programming courses such as
\href{https://github.com/ContextLab/storytelling-with-data}{\textit{Storytelling
    with Data}} \cite{Mann21b} (an open course on data science,
visualization, and communication) to simplify distributing lessons and
submitting assignments, as well as in online demos such as
{\texttt{abstract2paper}} \cite{Mann21a} (an example application of
\href{https://github.com/EleutherAI/gpt-neo}{GPT-Neo}) to share
ready-to-run code that installs dependencies automatically.

%\textbf{This is the main section of the article and the reviewers weight the description here appropriately}

%Indicate in what way new research questions can be pursued as a result of the software (if any).

%Indicate in what way, and to what extent, the pursuit of existing research questions is improved (if so).

%Indicate in what way the software has changed the daily practice of its users (if so).

%Indicate how widespread the use of the software is within and outside the intended user group.

%Indicate in what way the software is used in commercial settings and/or how it led to the creation of spin-off companies (if so).

@@

\section*{Author Contributions}
\textbf{Paxton C. Fitzpatrick}: Conceptualization, Methodology, Software, Validation, Writing - Original Draft, Visualization. \textbf{Jeremy R. Manning}: Conceptualization, Resources, Writing - Review \& Editing, Supervision, Funding acquisition.

\section*{Funding}
Our work was supported in part by NSF grant number 2145172 to J.R.M.
The content is solely the responsibility of the authors and does not necessarily represent the official views of our supporting organizations.

@@

\section*{Acknowledgements}

>>>>

Code sharing is a core component of the open science movement
that has inspired a full ecosystem of tools and packages.  However,
sharing code, in and of itself, does not guarantee that others will be
able to \textit{reproduce} the desired results.  For example, research
code often requires installing other software packages that extends
the language's basic functionality.  Within the Python community~\citep{vanR95},
external packages that are published in the most popular 
repositories~\citep{pypi, Anac12} are associated with version numbers
and tags that enable users to guarantee that they are installing
exactly the same code across different computing environments.
Despite that it is \textit{possible} to manually install the intended version numbers of
every dependency of a Python script or package, doing so may cause
conflicts within the user's computing environment that interferes with
the functionality of  \textit{other} code.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python
    ecosystem.}  From left to right: plain-text Python scripts (.py) provide the
  most basic system for sharing raw code.  Scripts may reference
  external libraries, but those libraries are not automatically
  installed on other users' systems, nor is any version checking
  performed by default.  Jupyter notebooks (.ipynb files) comprise embedded text,
  executable code, and media (including rendered figures, code output,
  etc.).  When the \texttt{davos} library is imported into a Jupyter
  notebook, the notebook's functionality is extended to automatically
  install the required external libraries (at their correct versions,
  when specified).  Virtual environments install an isolated copy of
  Python and all required dependencies.  This typically requires
  defining a \texttt{requirements.txt} file that lists all
  dependencies (including version numbers) along with an environment
  (.yml) file that specifies how the virtual environment should be
  configured.  Containers provide a means of defining an isolated
  environment that includes a complete operating system (independent
  of the user's operating system), in addition to a virtual
  environment or other configurations needed to provide the necessary
  computing environment.  Containers are typically defined using
  specification files (e.g., a plain-text \texttt{dockerfile}) that
  instruct the virtualization engine regarding how to build the
  virtual environment.  Virtual machines provide a complete
  hardware-level simulation of the computing environment.  In addition
  to simulating specific hardware, virtual machines (typically
  specified using binary images files) must also define operating
  system-level properties of the computing environment.  Systems to
  the left of the blue vertical line entail sharing using individual
  files, with no additional installation or configuration needed to
  run the target software.  Systems to the right of the red vertical
  line provide high reproducibility by supporting precise control over
  dependencies and versioning.}
\label{fig:code-sharing}
\end{figure}

@@

At another extreme, virtual machines~\citep{Gold74, AltiEtal05, vmware} provide a complete hardware-level
simulation of the desired system.  Virtual machines are typically
fully isolated from the user's system such that installing or running
software on a virtual machine does not impact the user's primary
operating system or computing environment.  Containers~\citep[e.g.,][]{Merk14,
  KurtEtal17} provide a similar ``isolated'' experience.  Although
containerized environments do not specify hardware-level operations,
they are typically packaged with a complete operating system, in
addition to a complete copy of Python and any relevant dependencies.
Virtual environments~\citep[e.g.,][]{vanREtal14} also provide a
computing environment that is largely separated from the user's main
environment.  They incorporate a copy of Python and the target
software's dependencies, but virtual environments do not specify or
reproduce a complete copy of the original operating system.  Each of
these systems (virtual machines, containers, and virtual environments)
guarantees (to differing degrees-- at the hardware-level, operating system
level, and Python environment, respectively) that the relevant code
will run similarly for different users.  However, each of these
systems also requires installing independent tools that
can be resource intensive or burdensome to install or configure.

We designed \texttt{davos} to occupy a ``sweet spot'' between these
extremes.  \texttt{davos} is a notebook-installable package that adds
functionality to the original.  Like standard Jupyter notebooks,
\texttt{davos}-enhanced notebooks allows researchers to include text,
executable code, and media within a single file.  No further setup or
installation is required, beyond what is needed to run standard
Jupyter notebooks.  And like virtual environments \texttt{davos}
provides a convenient mechanism for fully specifying (and installing,
as needed) a complete set of Python dependencies, including package versions.

@@

\subsection{Software architecture}
The \texttt{davos} package is structured as two sub-packages: a set of ``core" modules that implement...

@@

% ------------------------------------------------------- IMPACT -------------------------------------------------------
\section{Impact}

Like virtual environments, containers, and virtual machines, 
the \texttt{davos} library (when used in conjunction with Jupyter
notebook) provides a lightweight mechanism for sharing code and
ensuring reproducibility across users and computing environments.
Further,\texttt{davos} enables users to fully specify (and
install, as needed) any project dependencies within the same
notebook.  This provides a system whereby executable code (along with
text and media) \textit{and} code for setting up and configuring the
project dependencies, all within a single notebook file.

We designed \texttt{davos} for use in research applications.  For
example, in many cases \texttt{davos} may be used as a drop-in
replacement for more-difficult-to-set-up virtual environments,
containers, and/or virtual machines.  For researchers, this lowers barriers to sharing
code.  And by eliminating most of the setup costs of reconstructing
the researchers' computing environment, \texttt{davos} also serves to
lower barriers to entry for members of the scientific community and of
the public to \textit{benefiting} from shared code.

Beyond research applications, \texttt{davos} is also useful in
pedagogical settings.  For example, in programming courses,
instructors and students may choose to import the \texttt{davos}
library into their notebooks to provide a simple means of ensuring
their code will run on others' machines.  When combined with online
notebook-based platforms like Google Colaboratory, \texttt{davos}
provides a convenient means of managing dependencies within a
notebook, without requiring any software (beyond a web browser) to be
installed on the students' or instructors' systems.  For the same
reasons, \texttt{davos} also provides an elegant means of sharing
ready-to-run notebook-based demonstrations that install their
dependencies automatically.

In addition to the above common uses of \texttt{davos}, our work also
provides several more subtle ``advanced'' use cases.  
Whereas Python's built-in \texttt{import} statement is agnostic to
packages' version numbers, \texttt{smuggle} statements (when combined
with onion comments) are version-sensitive.  This setup enables
multiple versions of a single library to be imported within the same
notebook.  This could be useful in cases where specific features were
added or removed from a package across different versions, or in
comparing the performance or functionality of particular features
across different versions of the same package.

A second advanced use case is in providing a proof of concept of how
one can add new ``keywords'' to the Python language by leveraging the
error-handling mechanisms built into Jupyter notebooks.  This could
lead to exciting new tools that, like \texttt{davos}, extend the
Python language in useful ways.  We note that our approach to adding
the \texttt{smuggle} keyword to Python when \texttt{davos} is imported
into a notebook-based environment also carries some potential risk.
For example, a malicious user could use a similar approach to
substantially change a notebook's functionality, in
difficult-to-predict ways when a particular library was imported into
the current environment.  This highlights an important reason why
security-conscious users would be well-served to only make use of
libraries from trusted sources, or whose code is publicly available
for review.

@@

Perhaps the most exciting uses of the \texttt{davos} library are those
that we have \textit{not} yet considered or imagined.  We hope that
the Python community will find \texttt{davos} to provide a convenient
means of managing project dependencies to facilitate code sharing.  We
also hope that some of the more advanced applications of our library
might lead to new insights or discoveries.

\section*{Author Contributions}
Conceptualization: PCF and JRM.  Methodology: PCF and JRM.
Implementation: PCF.  Validation: PCF.  Testing: PCF and JRM.
Writing: PCF and JRM.

\section*{Funding}
Our work was supported in part by NSF grant number 2145172 to JRM.
The content is solely the responsibility of the authors and does not necessarily represent the official views of our supporting organizations.

@@

\section*{Acknowledgements}
We acknowledge useful feedback and discussion from the students of
JRM's \textit{Storytelling with Data} course (Winter, 2022 offering)
who used preliminary versions of our library in several assignments.
